### 2016-03-01 paper/icfp16-submission/semantics.tex
much cleaning and simplification
- simplify unification calls
- use \alpha instead of \boxdot for type holes
- pass \tsu to \gen so we don't have to apply it recursively in the call
- write `leaf[t]` instead of `leaf @t`, looks better in tex
- add a bunch of ES notes
- extend proof appendix

\subsection{Syntax}
\label{sec:syntax}
\input{syntax.tex}
%
Figure~\ref{fig:syntax} describes the syntax of \lang, a simple lambda
calculus with integers, booleans, and binary trees.
%
As we are specifically interested in programs that \emph{do} go wrong,
we include an explicit \stuck\ configuration in our syntax.

\paragraph{Holes}
\label{sec:holes}
%
Recall that a key challenge in our setting is to find witnesses
that are meaningful and do not arise from choosing values from
irrelevant types.
%
We solve this problem be equipping our term language with a notion
of a \emph{hole}, written \ehole, which represents \emph{unconstrained}
values that may be replaced with values from \emph{any} type.
%
Dually, holes may also appear in types, written \thole, where they may
be thought of as type variables that we will not generalize over.
%

@@

\paragraph{Narrowing Types} The procedure % $\force{v}{t}{\vsu}{\tsu}$,
$$
\forcesym :: (v, t, \vsu, \tsu) \rightarrow \triple{v \cup \stuck}{\vsu}{\tsu}
$$
defined in Figure~\ref{fig:narrow}, takes as input a value $v$, a type
$t$, and the current value and type substitutions, and refines $v$ to
have type $t$ by yielding a triple of either the same value and
substitutions, or yields the stuck state if no such refinement is
possible. In the case where $v$ is a hole, it first checks in the given
$\vsu$ to see if the hole has already been instantiated and, if so,
returns the existing instantiation.
\ES{explain the subtlety of narrowing a value hole to an uninstantiated type}
%
\begin{figure*}[ht]
$$
\begin{array}{lcl}
%% \multicolumn{3}{l}{\forcesym \ :: \  (e, t) \rightarrow \pair{e}{\vsu}} \\
\forcesym                  & ::     & (v, t, \vsu, \tsu) \rightarrow \triple{v \cup \stuck}{\vsu}{\tsu} \\
\force{v}{\thole}{\vsu}{\tsu}  & \defeq & \hspace{-1ex}
\begin{cases}
  \force{v}{\subst{\tsu}{\thole}}{\vsu}{\tsu} & \mbox{if}\ \thole \in dom(\tsu) \\
  \triple{v}{\vsu}{\extendsu{\tsu}{\thole}{\typeof{v}}} & \mbox{otherwise} \\
\end{cases} \\
\force{\vhole{t_1}}{t_2}{\vsu}{\tsu} & \defeq & \hspace{-1ex}
\begin{cases}
  \triple{v}{\vsu}{\tsu''}    & \mbox{if } v = \lookupsu{\vsu}{\ehole},
                                         \tsu' = \unify{t_1}{t_2}{\tsu},
                                         \tsu'' = \unify{\subst{\tsu'}{t_1}}
                                                        {\typeof{v}}{\tsu'} \\
  \triple{\stuck}{\vsu}{\tsu} & \mbox{if } v = \lookupsu{\vsu}{\ehole} \\
  \triple{v}{\extendsu{\vsu}{\ehole}{v}}{\tsu'} & \mbox{if}\ \tsu' = \unify{t_1}{t_2}{\tsu}, v = \gen{\subst{\tsu'}{t_1}}\\
\end{cases} \\
% \begin{cases}
%   \triple{\lookupsu{\vsu}{\ehole}}{\vsu}{\tsu} & \mbox{if}\ \ehole \in dom(\vsu), \hastype{\lookupsu{\vsu}{\ehole}}{t} \\
%   \triple{\stuck}{\vsu}{\tsu}                  & \mbox{if}\ \ehole \in dom(\vsu) \\
%   \triple{v}{\extendsu{\vsu}{\ehole}{v}}{\tsu}   & v = \gen{t}\\
% \end{cases} \\
 % \begin{array}{l}
    % \mathtt{if}\ i \in \vsu\ \mathtt{then}\ \triple{\vsu(i)}{\vsu}\ \mathtt{else} \\
         % \elet{v}{\gen{t}}{\triple{v}{\ehole{i} \mapsto v}}
  % \end{array} \\
\force{n}{\tint}{\vsu}{\tsu}     & \defeq & \triple{n}{\vsu}{\tsu} \\
\force{b}{\tbool}{\vsu}{\tsu}    & \defeq & \triple{b}{\vsu}{\tsu} \\
\force{\efun{x}{e}}{\tfun}{\vsu}{\tsu} & \defeq & \triple{\efun{x}{e}}{\vsu}{\tsu} \\
\force{\vleaf{t_1}}{\ttree{t_2}}{\vsu}{\tsu} & \defeq & \triple{\vleaf{t_1}}{\vsu}{\tsu'}, \mbox{if}\ \tsu' = \unify{t_1}{t_2}{\tsu} \\
\force{\vnode{t_1}{v_1}{v_2}{v_3}}{\ttree{t_2}}{\vsu}{\tsu} & \defeq & \triple{\vnode{t_1}{v_1}{v_2}{v_3}}{\vsu}{\tsu'}, \mbox{if}\ \tsu' = \unify{t_1}{t_2}{\tsu} \\
\force{v}{t}{\vsu}{\tsu} & \defeq & \triple{\stuck}{\vsu}{\tsu}
\end{array}
$$
\caption{Narrowing values}
\label{fig:narrow}
\end{figure*}
%
While a hole may map to a value that \emph{contains} another hole, \eg a
lambda or a tree, it may not map \emph{directly} to another hole, so in
the second case of \forcesym\ we do not need to \forcesym\ the result of the
substitution, the sub-hole will be narrowed when the context demands it.

\paragraph{Generating Values} The (non-deterministic) procedure $\gen{t}$
takes as input a type $t$ and returns a value of that type. For base
types, the procedure returns an arbitrary value. For functions, it returns
a lambda with a \emph{new} hole denoting the return value, and unconstrained
types (denoted by $\thole$) yield fresh, unconstrained values
(denoted by $\ehole$).

$$
\begin{array}{lcll}
\gensym       & ::      & t \rightarrow v \\
\gen{\tint}   & \defeq  & n, & \quad \text{non-deterministic} \\
\gen{\tbool}  & \defeq  & b, & \quad \text{non-deterministic} \\
\gen{\ttree{t}}  & \defeq  & tr, & \quad \text{non-deterministic, well-typed} \\
\gen{\tfun}   & \defeq & \efun{x}{\ehole}, & \quad \text{\ehole is fresh} \\
\gen{\thole}  & \defeq & \vhole{\thole}, & \quad \text{\ehole is fresh} \\
\end{array}
$$

\paragraph{Steps and Traces}
\input{operational.tex}
%
% WRW notes that Figure 4 does not seem to handle recursion (it's not clear
% how the let rule would work for something "let rec"-y, and there's not
% function call rule). I only mention this because I can imagine a reviewer
% wondering about your ability to generate good witnesses for function
% types. This could likely be addressed in text, by a forward reference to
% Section 3.4 where higher-order functions are handled, without changing
% any of the formalisms at the last minute.
Figure~\ref{fig:operational} describes the small-step contextual
reduction semantics for \lang.
%
A \emph{configuration} is a triple $\triple{e \cup \stuck}{\vsu}{\tsu}$
of an expression $e$ or the stuck term $\stuck$, a value substitution
$\vsu$, and a type substitution $\tsu$.
%
We write $\step{e}{\vsu}{\tsu}{e'}{\vsu'}{\tsu'}$ if the configuration
$\triple{e}{\vsu}{\tsu}$ transitions in a \emph{small step} to
$\triple{e'}{\vsu'}{\tsu'}$.
%
A (finite) \emph{trace} $\trace$ is a sequence of configurations
$\triple{e_0}{\vsu_0}{\tsu_0}, \ldots, \triple{e_n}{\vsu_n}{\tsu_n}$ such that
$\forall 0 \leq i < n$, we have
$\step{e_i}{\vsu_i}{\tsu_i}{e_{i+1}}{\vsu_{i+1}}{\tsu_{i+1}}$.
%
We write \steptr{\trace}{e}{\vsu}{\tsu}{e'}{\vsu'}{\tsu'} if $\trace$ is
a trace of the form $\triple{e}{\vsu}{\tsu},\ldots,\triple{e'}{\vsu'}{\tsu'}$.
%
We write \steps{e}{\vsu}{\tsu}{e'}{\vsu'}{\tsu'} if
\steptr{\trace}{e}{\vsu}{\tsu}{e'}{\vsu'}{\tsu'} for some trace $\trace$.

@@

\paragraph{Witness Generality}
We now show that our evaluation relation instantiates holes in a
\emph{general} manner. That is, given a function $f$, if we have
$\steps{\eapp{f}{\ehole{}}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}$,
then for every possible input type $t$, we can find a value
$v$ of type $t$ such that $\eapp{f}{v}$ goes wrong.

\begin{thm}{\textbf{[Witness Generality]}}
\label{thm:soundness}
  For any function $f$, if\\
  \hbox{$\steptr{\trace}{\eapp{f}{\ehole}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}$,}
  then for every % inhabitable
  type
  % \footnote{We exclude builtin functions that subvert the type system, \eg \texttt{Obj.magic}, and thus consider the type $\forall a b. \tfun{a}{b}$ to be uninhabitable.}
  $t$ there exists $\hastype{v}{t}$ such that
  $\steps{\eapp{f}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}$.
\end{thm}

@@

\paragraph{The Type of an Expression} The (dynamic) type of an
expression $e$ is defined as a function $\typeof{e}$ defined as:
  \[
  \begin{array}{lcll}
    \typeof{n}   & \defeq & \tint & \\
    \typeof{b}   & \defeq & \tbool & \\
    \typeof{\efun{x}{e}} & \defeq & \tfun \\
    \typeof{\vleaf{t}} & \defeq & \ttree{t} \\
    \typeof{\vnode{t}{v_1}{v_2}{v_3}} & \defeq & \ttree{t} \\
    \typeof{\vhole{t}} & \defeq & t \\
    \typeof{\eleaf} & \defeq & \ttree{\thole}, & \quad \text{\thole is fresh} \\
    \typeof{\enode{v_1}{v_2}{v_3}} & \defeq & \ttree{\thole}, & \quad \text{\thole is fresh} \\
    \typeof{e} & \defeq & \thole, & \quad \text{\thole is fresh} \\
  \end{array}
  \]
%
Note that we approximate the type of expressions rather than
performing full type inference, for example
$\typeof{\efun{x}{\eplus{x}{1}}} = \tfun$,
instead of the usual $\tint \rightarrow \tint$.

\paragraph{Dynamic Type Inference}
We can think of the evaluation of \eapp{f}{\ehole} as (dynamically)
inferring a \emph{partial type} --- a type that may contain holes ---
for $f$'s input.
%
We can extract this type from evaluation traces.
%
Formally, we say that if
$\steptr{\trace}{\eapp{f}{\ehole{}}}{\emptysu}{\emptysu}{e}{\vsu}{\tsu}$,
then the \emph{partial input type} of $f$ upto $\trace$, written
\ptype{\trace}{f}, is
$\typeof{\subst{\su}{\ehole{}}}$.
%
%We will omit the subscript when we wish to refer to the final partial
%type, \ie\ at the step where the expression has been reduced to a value
%(or stuck.)

@@

\paragraph{Preservation}
Given these two definitions, we show that each evaluation step
refines the partial input type of $f$, \ie preserves type compatibility.
%
\begin{lem}
\label{lem:refine-partial}
If $\trace \defeq \triple{\eapp{f}{\ehole{}}}{\emptysu}{\emptysu},\ldots$ and
$\trace' \defeq \trace, \triple{e'}{\vsu'}{\tsu'}$ (\ie $\trace'$ is a single-step
extension of $\trace$)
%
%The partial type of $f$ upto $\trace$ is compatible
%with the partial type upto $\trace'$, \ie\
%
then \tcompat{\ptype{\trace}{f}}{\ptype{\trace'}{f}}.
\end{lem}
\begin{proof}
  By case analysis on the evaluation rules.
  %
  Note that all rules preserve partial types with the exception of when
  \forcesym\ is called on a hole, in which case we instantiate the hole with
  a concrete value.
  %
  But $\typeof{\ehole{}} = \thole{}$, which is compatible with any type.
\end{proof}

\paragraph{Narrowing}
%
Furthermore, only a successful call to \forcesym can change the partial type of $f$.
%
\begin{lem}
\label{lem:force-inst}
If
$\trace \defeq \triple{\eapp{f}{\ehole{}}}{\emptysu}{\emptysu},\ldots,\triple{e}{\vsu}{\tsu}$
and
$\trace' \defeq \trace, \step{e}{\vsu}{\tsu}{e'}{\vsu'}{\tsu'}$
and
$\ptype{\trace}{f} \neq \ptype{\trace'}{f}$,
then the final step $\step{e}{\vsu}{\tsu}{e'}{\vsu'}{\tsu'}$ invokes \forcesym.
\end{lem}

@@

\paragraph{Incompatible Values Are Wrong}
%
\emph{Any} value that is \emph{incompatible} with
the partial type upto trace $\trace$ will cause $f$ to get stuck
in \emph{at most} $k$ steps, where $k$ is the length of $\trace$.
%
\begin{lem}
\label{lem:k-stuck}
  If \steptr{\trace}{\eapp{f}{\ehole}}{\emptysu}{\emptysu}{e}{\vsu}{\tsu} and
     \hastype{v}{t} and
     \tincompat{t}{\ptype{\trace}{f}},
  then
     \steps{\eapp{f}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}.
\end{lem}
\begin{proof}
  By induction on $k$, the length of $\trace$.
  Let $s_{\trace} = \ptype{\trace}{f}$ and
  $\trace' = \triple{\eapp{f}{\ehole{}}}{\emptysu}{\emptysu},...,\triple{e'}{\vsu'}{\tsu'}$
  such that $\step{e'}{\vsu'}{\tsu'}{e}{\vsu}{\tsu}$.
  Suppose \hastype{v}{\tincompat{t}{s_{\trace}}}, we
  will show that \steps{\eapp{f}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu'}{\tsu'}
  in at most $k$ steps:
%
  \begin{description}
  \item [Case \tincompat{t}{s_{\trace'}}:]
    The inductive hypothesis applies.
  \item [Case $\tcompat{t}{s_{\trace'}}$ but $\tincompat{t}{s_{\trace}}$:]
    By Lemma~\ref{lem:force-inst} we must have called \forcesym at step
    $k$.
    %
    Lemma~\ref{lem:refine-partial} tells us
    $\tcompat{s_{\trace'}}{s_{\trace}}$, which means we must have
    specifically narrowed $s_{\trace'}$ to a type incompatible with $t$.
    %
    A case analysis of the rules shows that \forcesym cannot have
    succeeded.
  \end{description}
\end{proof}

\begin{proof}[\textbf{Proof of Theorem~\ref{thm:soundness}}]
%
Suppose $\trace$ witnesses that $f$ gets stuck,
and let $t = \ptype{\trace}{f}$.
We show that \emph{all} types $s$ have stuck-inducing
values by splitting cases on whether the type is
compatible with $t$. %the partial type upto $\trace$.
%
\begin{description}
\item [Case \tcompat{s}{t}:]
  Let $\trace = \triple{\eapp{f}{\ehole{}}}{\emptysu}{\emptysu},...,\triple{\stuck}{\vsu}{\tsu}$.
  %
  The value $v = \subst{\vsu}{\ehole}$ demonstrates that
  $\eapp{f}{v}$ gets stuck.
\item [Case \tincompat{s}{t}:] By Lemma~\ref{lem:k-stuck}, every $v$
  such that \hastype{v}{t} demonstrates that $\eapp{f}{v}$ gets stuck.
\end{description}
\end{proof}

@@

\begin{figure}[t]
\centering
\begin{mcode}
genArgs :: [($x$, $e$)] -> [$v$]
genArgs bs = loop []
  where
  loop vs  = case eval (close bs vs) of
               $\pair{\efun{x}{e}}{\_}$:_ -> loop ($\ehole{}$:vs)
               _        -> vs

@@

\begin{cor}{\textbf{[Witness Generation]}}
\label{thm:generation}
  If @genWitness k [(f1,e1),...,(fn,en)]@ is non-empty,
  then for every type $t$ there exists $\hastype{v}{t}$ such that
  $\steps{\eapp{f_n}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}$.
\end{cor}

>>>>

\subsection{Syntax}
\label{sec:syntax}
\input{syntax.tex}
%
Figure~\ref{fig:syntax} describes the syntax of \lang, a simple lambda
calculus with integers, booleans, and binary trees.
%
As we are specifically interested in programs that \emph{do} go wrong,
we include an explicit \stuck\ state in our syntax.

\paragraph{Holes}
\label{sec:holes}
%
Recall that a key challenge in our setting is to find witnesses
that are meaningful and do not arise from choosing values from
irrelevant types.
%
We solve this problem be equipping our term language with a notion
of a \emph{hole}, written \vhole{\thole}, which represents an \emph{unconstrained}
value that may be replaced with any value of an unspecified type \thole.
%
Dually, holes may also appear in types, written \thole, where they may
be thought of as type variables that we will not generalize over.
%

@@

\paragraph{Narrowing Types} The procedure % $\force{v}{t}{\vsu}{\tsu}$,
$$
\forcesym :: (v, t, \vsu, \tsu) \rightarrow \triple{v \cup \stuck}{\vsu}{\tsu}
$$
defined in Figure~\ref{fig:narrow}, takes as input a value $v$, a type
$t$, and the current value and type substitutions, and refines $v$ to
have type $t$ by yielding a triple of either the same value and
substitutions, or yields the stuck state if no such refinement is
possible. In the case where $v$ is a hole, it first checks in the given
$\vsu$ to see if the hole has already been instantiated and, if so,
returns the existing instantiation.
\ES{explain the subtlety of narrowing a value hole to an uninstantiated type}
%
\begin{figure*}[ht]
$$
\begin{array}{lcl}
%% \multicolumn{3}{l}{\forcesym \ :: \  (e, t) \rightarrow \pair{e}{\vsu}} \\
\forcesym                  & ::     & (v, t, \vsu, \tsu) \rightarrow \triple{v \cup \stuck}{\vsu}{\tsu} \\
\force{v}{\thole}{\vsu}{\tsu}  & \defeq & \hspace{-1ex}
\begin{cases}
  \triple{v}{\vsu}{\tsu'} & \mbox{if } \tsu' = \unify{\{\typeof{v}, \thole\}}{\tsu} \\
  \triple{\stuck}{\vsu}{\tsu} & \mbox{otherwise} \\

  % \force{v}{\subst{\tsu}{\thole}}{\vsu}{\tsu} & \mbox{if}\ \thole \in dom(\tsu) \\
  % \triple{v}{\vsu}{\extendsu{\tsu}{\thole}{\typeof{v}}} & \mbox{otherwise} \\
\end{cases} \\
\force{\vhole{t_1}}{t_2}{\vsu}{\tsu} & \defeq & \hspace{-1ex}
\begin{cases}
  \triple{v}{\vsu}{\tsu'}    & \mbox{if } v = \lookupsu{\vsu}{\ehole},
                                         \tsu' = \unify{\{t_1, t_2, \typeof{v}\}}{\tsu}\\
  \triple{\stuck}{\vsu}{\tsu} & \mbox{if } v = \lookupsu{\vsu}{\ehole} \\
  \triple{v}{\extendsu{\vsu}{\ehole}{v}}{\tsu'} & \mbox{if}\ \tsu' = \unify{\{t_1, t_2\}}{\tsu}, v = \gen{t_1}{\tsu'}\\
\end{cases} \\
% \begin{cases}
%   \triple{\lookupsu{\vsu}{\ehole}}{\vsu}{\tsu} & \mbox{if}\ \ehole \in dom(\vsu), \hastype{\lookupsu{\vsu}{\ehole}}{t} \\
%   \triple{\stuck}{\vsu}{\tsu}                  & \mbox{if}\ \ehole \in dom(\vsu) \\
%   \triple{v}{\extendsu{\vsu}{\ehole}{v}}{\tsu}   & v = \gen{t}\\
% \end{cases} \\
 % \begin{array}{l}
    % \mathtt{if}\ i \in \vsu\ \mathtt{then}\ \triple{\vsu(i)}{\vsu}\ \mathtt{else} \\
         % \elet{v}{\gen{t}}{\triple{v}{\ehole{i} \mapsto v}}
  % \end{array} \\
\force{n}{\tint}{\vsu}{\tsu}     & \defeq & \triple{n}{\vsu}{\tsu} \\
\force{b}{\tbool}{\vsu}{\tsu}    & \defeq & \triple{b}{\vsu}{\tsu} \\
\force{\efun{x}{e}}{\tfun}{\vsu}{\tsu} & \defeq & \triple{\efun{x}{e}}{\vsu}{\tsu} \\
\force{\vleaf{t_1}}{\ttree{t_2}}{\vsu}{\tsu} & \defeq & \triple{\vleaf{t_1}}{\vsu}{\tsu'}, \mbox{if}\ \tsu' = \unify{\{t_1, t_2\}}{\tsu} \\
\force{\vnode{t_1}{v_1}{v_2}{v_3}}{\ttree{t_2}}{\vsu}{\tsu} & \defeq & \triple{\vnode{t_1}{v_1}{v_2}{v_3}}{\vsu}{\tsu'}, \mbox{if}\ \tsu' = \unify{\{t_1, t_2\}}{\tsu} \\
\force{v}{t}{\vsu}{\tsu} & \defeq & \triple{\stuck}{\vsu}{\tsu}
\end{array}
$$
\caption{Narrowing values}
\label{fig:narrow}
\end{figure*}
%
While a hole may map to a value that \emph{contains} another hole, \eg a
lambda or a tree, it may not map \emph{directly} to another hole, so in
the second case of \forcesym\ we do not need to \forcesym\ the result of the
substitution, the sub-hole will be narrowed when the context demands it.

\paragraph{Generating Values} The (non-deterministic) procedure $\gen{t}$
takes as input a type $t$ and returns a value of that type. For base
types, the procedure returns an arbitrary value. For functions, it returns
a lambda with a \emph{new} hole denoting the return value, and unconstrained
types (denoted by $\thole$) yield fresh holes constrained to have type \thole
(denoted by $\vhole{\thole}$). \ES{note that $\gen{\ttree{t}}{\tsu}$ will
always place holes in the generated nodes, they will be filled in later,
on demand.}

$$
\begin{array}{lcll}
\gensym       & ::  & (t, \tsu) \rightarrow v \\
\gen{\thole}{\tsu}  & \defeq  & \gen{\subst{\tsu}{\thole}}{\tsu}, & \quad \text{if } \thole \in dom(\tsu) \\
\gen{\tint}{\tsu}   & \defeq  & n, & \quad \text{non-deterministic} \\
\gen{\tbool}{\tsu}  & \defeq  & b, & \quad \text{non-deterministic} \\
\gen{\ttree{t}}{\tsu}  & \defeq  & tr, & \quad \text{non-deterministic} \\
\gen{\tfun}{\tsu}   & \defeq & \efun{x}{\vhole{\thole}}, & \quad \text{\ehole and \thole are fresh} \\
\gen{\thole}{\tsu}  & \defeq & \vhole{\thole}, & \quad \text{\ehole is fresh} \\
\end{array}
$$

\paragraph{Steps and Traces}
\input{operational.tex}
%
% WRW notes that Figure 4 does not seem to handle recursion (it's not clear
% how the let rule would work for something "let rec"-y, and there's not
% function call rule). I only mention this because I can imagine a reviewer
% wondering about your ability to generate good witnesses for function
% types. This could likely be addressed in text, by a forward reference to
% Section 3.4 where higher-order functions are handled, without changing
% any of the formalisms at the last minute.
Figure~\ref{fig:operational} describes the small-step contextual
reduction semantics for \lang.
%
A program state is a triple $\triple{e \cup \stuck}{\vsu}{\tsu}$ of an
expression $e$ or the stuck term $\stuck$, a value substitution $\vsu$,
and a type substitution $\tsu$. In the sequel we will write \estuck to
denote either an expression $e$ or \stuck.
%
We write $\step{\estuck}{\vsu}{\tsu}{\estuck'}{\vsu'}{\tsu'}$ if the state
$\triple{\estuck}{\vsu}{\tsu}$ transitions in a \emph{single step} to
$\triple{\estuck'}{\vsu'}{\tsu'}$.
%
A (finite) \emph{trace} $\trace$ is a sequence of configurations
$\triple{\estuck_0}{\vsu_0}{\tsu_0}, \ldots, \triple{\estuck_n}{\vsu_n}{\tsu_n}$ such that
$\forall 0 \leq i < n$, we have
$\step{\estuck_i}{\vsu_i}{\tsu_i}{\estuck_{i+1}}{\vsu_{i+1}}{\tsu_{i+1}}$.
%
We write \steptr{\trace}{\estuck}{\vsu}{\tsu}{\estuck'}{\vsu'}{\tsu'} if $\trace$ is
a trace of the form $\triple{\estuck}{\vsu}{\tsu},\ldots,\triple{\estuck'}{\vsu'}{\tsu'}$.
%
We write \steps{\estuck}{\vsu}{\tsu}{\estuck'}{\vsu'}{\tsu'} if
\steptr{\trace}{\estuck}{\vsu}{\tsu}{\estuck'}{\vsu'}{\tsu'} for some trace $\trace$.

@@

\paragraph{Witness Generality}
We now show that our evaluation relation instantiates holes in a
\emph{general} manner. That is, given a function $f$, if we have
$\steps{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}$,
then for every possible input type $t$, we can find a value
$v$ of type $t$ such that $\eapp{f}{v}$ goes wrong.

\begin{thm}{\textbf{[Witness Generality]}}
\label{thm:soundness}
  For any function $f$, if\\
  \hbox{$\steptr{\trace}{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}$,}
  then for every % inhabitable
  type
  % \footnote{We exclude builtin functions that subvert the type system, \eg \texttt{Obj.magic}, and thus consider the type $\forall a b. \tfun{a}{b}$ to be uninhabitable.}
  $t$ there exists $\hastype{v}{t}$ such that
  $\steps{\eapp{f}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}$.
\end{thm}

@@

\paragraph{The Type of an Expression} The (dynamic) type of an
expression $e$ is defined as a function $\typeof{e}$ defined as:
  \[
  \begin{array}{lcll}
    \typeof{n}   & \defeq & \tint & \\
    \typeof{b}   & \defeq & \tbool & \\
    \typeof{\efun{x}{e}} & \defeq & \tfun \\
    \typeof{\vleaf{t}} & \defeq & \ttree{t} \\
    \typeof{\vnode{t}{v_1}{v_2}{v_3}} & \defeq & \ttree{t} \\
    \typeof{\vhole{\thole}} & \defeq & \thole \\
    \typeof{\eleaf} & \defeq & \ttree{\thole}, & \quad \text{\thole is fresh} \\
    \typeof{\enode{v_1}{v_2}{v_3}} & \defeq & \ttree{\thole}, & \quad \text{\thole is fresh} \\
    \typeof{e} & \defeq & \thole, & \quad \text{\thole is fresh} \\
  \end{array}
  \]
%
Note that we approximate the type of expressions rather than
performing full type inference, for example
$\typeof{\efun{x}{\eplus{x}{1}}} = \tfun$,
instead of the usual $\tint \rightarrow \tint$.

\paragraph{Dynamic Type Inference}
We can think of the evaluation of \eapp{f}{\vhole{\thole}} as (dynamically)
inferring a \emph{partial type} --- a type that may contain holes ---
for $f$'s input, \ie a partial instantiation of \thole.
%
We can extract this type from an evaluation trace.
%
Formally, we say that if
$\steptr{\trace}{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}{\estuck}{\vsu}{\tsu}$,
then the \emph{partial input type} of $f$ upto $\trace$, written
\ptype{\trace}{f}, is
$\typeof{\subst{\vsu}{\ehole}}$.
\ES{should we say $\subst{\tsu}{\thole}$ instead? would need a helper function that does the repeated application of \tsu\ until the result has no $\thole \in dom(\tsu)$}
%
%We will omit the subscript when we wish to refer to the final partial
%type, \ie\ at the step where the expression has been reduced to a value
%(or stuck.)

@@

\paragraph{Preservation}
Given these two definitions, we show that each evaluation step
refines the partial input type of $f$, \ie preserves type compatibility.
%
\begin{lem}
\label{lem:refine-partial}
If $\trace \defeq \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots$ and
$\trace' \defeq \trace, \triple{\estuck'}{\vsu'}{\tsu'}$ (\ie $\trace'$ is a single-step
extension of $\trace$)
%
%The partial type of $f$ upto $\trace$ is compatible
%with the partial type upto $\trace'$, \ie\
%
then \tcompat{\ptype{\trace}{f}}{\ptype{\trace'}{f}}.
\end{lem}
\begin{proof}
  By case analysis on the evaluation rules.
  %
  Note that all rules preserve partial types with the exception of when
  \forcesym\ is called on a hole, in which case we may instantiate the hole with
  a concrete value.
  %
  But $\typeof{\vhole{\thole}} = \thole$, which is compatible with any type.
\end{proof}

\paragraph{Narrowing}
%
Furthermore, only a successful call to \forcesym can change the partial type of $f$.
%
\begin{lem}
\label{lem:force-inst}
If
$\trace \defeq \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots,\triple{e}{\vsu}{\tsu}$
and
$\trace' \defeq \trace, \step{e}{\vsu}{\tsu}{\estuck'}{\vsu'}{\tsu'}$
and
$\ptype{\trace}{f} \neq \ptype{\trace'}{f}$,
then the final step $\step{e}{\vsu}{\tsu}{\estuck'}{\vsu'}{\tsu'}$ invokes \forcesym.
\end{lem}

@@

\paragraph{Incompatible Values Are Wrong}
%
\emph{Any} value that is \emph{incompatible} with
the partial type upto trace $\trace$ will cause $f$ to get stuck
in \emph{at most} $k$ steps, where $k$ is the length of $\trace$.
%
\begin{lem}
\label{lem:k-stuck}
  If \steptr{\trace}{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}{e}{\vsu}{\tsu} and
     \hastype{v}{t} and
     \tincompat{t}{\ptype{\trace}{f}},
  then
     \steps{\eapp{f}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}.
\end{lem}
\begin{proof}
  By induction on $k$, the length of $\trace$.
  Let $s_{\trace} = \ptype{\trace}{f}$ and
  $\trace' = \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},...,\triple{e'}{\vsu'}{\tsu'}$
  such that $\step{e'}{\vsu'}{\tsu'}{e}{\vsu}{\tsu}$.
  Suppose \hastype{v}{\tincompat{t}{s_{\trace}}}, we
  will show that \steps{\eapp{f}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu'}{\tsu'}
  in at most $k$ steps:
%
  \begin{description}
  \item [Case \tincompat{t}{s_{\trace'}}:]
    The inductive hypothesis applies.
  \item [Case $\tcompat{t}{s_{\trace'}}$ but $\tincompat{t}{s_{\trace}}$:]
    By Lemma~\ref{lem:force-inst} we must have called \forcesym at step
    $k$.
    %
    Lemma~\ref{lem:refine-partial} tells us
    $\tcompat{s_{\trace'}}{s_{\trace}}$, which means we must have
    specifically narrowed $s_{\trace'}$ to a type incompatible with $t$.
    %
    A case analysis of the rules shows that \forcesym cannot have
    succeeded.
  \end{description}
\end{proof}

\begin{proof}[\textbf{Proof of Theorem~\ref{thm:soundness}}]
%
Suppose $\trace$ witnesses that $f$ gets stuck,
and let $t = \ptype{\trace}{f}$.
We show that \emph{all} types $s$ have stuck-inducing
values by splitting cases on whether the type is
compatible with $t$. %the partial type upto $\trace$.
%
\begin{description}
\item [Case \tcompat{s}{t}:]
  Let $\trace = \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},...,\triple{\stuck}{\vsu}{\tsu}$.
  %
  The value $v = \subst{\vsu}{\ehole}$ demonstrates that
  $\eapp{f}{v}$ gets stuck. \ES{here we also need a repeated application of the substitution, as \ehole may map directly to $\vnode{\thole}{\vhole{\thole}}{\vleaf{\thole}}{\vleaf{\thole}}$}
\item [Case \tincompat{s}{t}:] By Lemma~\ref{lem:k-stuck}, every $v$
  such that \hastype{v}{t} demonstrates that $\eapp{f}{v}$ gets stuck.
\end{description}
\end{proof}

@@

\begin{figure}[t]
\centering
\begin{mcode}
genArgs :: [($x$, $e$)] -> [$v$]
genArgs bs = loop []
  where
  loop vs  = case eval (close bs vs) of
               $\pair{\efun{x}{e}}{\_}$:_ -> loop ($\vhole{\thole}$ : vs)
               _        -> vs

@@

\begin{cor}{\textbf{[Witness Generation]}}
\label{thm:generation}
  If @genWitness k [(f1,e1),...,(fn,en)]@ is non-empty,
  then for every type $t$ there exists $v$ of type $t$ such that
  $\steps{\eapp{f_n}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}$.
\end{cor}
